---
title: "Ocean Man; Marsh Population Regression"
author: "Jared Lourie"
date: "12/21/2020"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(car)
library(fitdistrplus)
library(emmeans)
library(betareg)
nektonxmacro <- read_csv('../data/nektonxmacro.csv')
nektonxmacro$creek <- factor(nektonxmacro$creek, levels = c('Clubhead', 'Sweeney', 'Nelson', 'West'))
nektonxmacro <- nektonxmacro %>% mutate(branch = as.factor(branch), type = as.factor(type), treatment = as.factor(treatment))
```
|    In order to understand the PIE LTER, you have to think like the Marsh Man or think like the green guy in the swamp, in tune with the environment and knees deep in the action as an ecologist should be. My first thoughts when approaching the ideas surrounding the TIDE project and understanding the impact of agricultural runoff were about ecological recovery from within the Plum Island ecosystem. If there were effects on the population or the geomorphology that would be adverse to those populations, would competitions and top-down or bottom-up effects end up reaching some sort of new stability, or would the ecosystem ultimately crash? Presumably a Pavlovian carrot thrown into the mix of a delicate ecosystem won't end over very well ([EPA facts on poor fertilizer control](https://www.epa.gov/nutrientpollution/sources-and-solutions-agriculture)). With this consideration in mind I joined data detailing information concerning Nekton and Macroinfaunal populations. For more context, the PIE LTER is the Plum Island Estuary long term ecology research group, a collaboration involving many professors and universities. The summary of the following merged dataset is created from these datasets created and uploaded by PI [Linda Deegan et. al](https://thetideproject.org/about/people/); [Macroinfaunal](https://pie-lter.ecosystems.mbl.edu/content/macroinfaunal-counts-2003-2013-marsh-cores-along-rowley-river-tidal-creeks-associated-long-t) & [Nekton](https://pie-lter.ecosystems.mbl.edu/content/nekton-species-counts-and-density-flume-net-collections-along-rowley-river-tidal-creeks-asso). 
```{r peak, echo=FALSE}
summary(nektonxmacro)
```
 
|    Looking into the data a bit and using the package [vegan](https://cran.r-project.org/web/packages/vegan/vegan.pdf) to assign [Shannon Diversity Index](https://stats.libretexts.org/Bookshelves/Applied_Statistics/Book%3A_Natural_Resources_Biometrics_(Kiernan)/10%3A_Quantitative_Measures_of_Diversity_Site_Similarity_and_Habitat_Suitability/10.01%3A_Introduction__Simpsons_Index_and_Shannon-Weiner_Index) values, an understanding of ecosystem eveness and abundance can be visualized.

```{r shannon_viz, echo=FALSE, warning=FALSE}
nektonxmacro %>% filter(month %in% c(7, 8, 9)) %>% ggplot(aes(x = date, y =  shannon, color = creek)) +
  geom_point() +
  stat_smooth(formula = 'y ~ x', method = 'loess') +
  facet_wrap(type~treatment) +
  theme_minimal() +
  scale_color_viridis_d()
```

|    Macro Control appears to have normal oscillations of Shannon Diversity eveness, indicative of a pattern of population equilibrium and likely a Lotka-Volterra zero growth [predator/prey](http://www.tiem.utk.edu/~gross/bioed/bealsmodules/predator-prey.html) & [competition](http://www.tiem.utk.edu/~gross/bioed/bealsmodules/competition.html) interaction. Interestingly enough nekton control seems to lose it's normal oscillation of diversity eveness, with both creeks tapering off and losing eveness and likely their nekton equilibria. Fertilized macro and nekton diversity curves appear to be tapering off with normal oscillation, indicative of a linear diversity or equilibria loss. A prelminary look at change in percent representation of species of interest by creek is not very revealing especially with some macro and nekton species being in the same model. As it appears above it seems as though nekton and macro overall are not linked. I would hypothesize that with the increased ability of aquatic species to travel throughout the PIE system, that nekton effects on population may spread, but without data or confirmation of individuals moving from treated to control creeks, it is only a thought. It is also entirely possible that the effects on nekton biodiversity are cause by an anthropogenic factor or other effect since the effect is prevalent in the control and treatment creeks.

|    According to [Sinéad M. Crotty et al](https://onlinelibrary-wiley-com.ezproxy.lib.umb.edu/doi/full/10.1111/ele.13146), salt marsh biodiveristy is affected mostly by foundation or keystone species which allow for biodiversity to grow and accumulate within a marsh patchwork. I hypothesize that coastal eutrophication of the PIE network [(Linda A. Deegan et. al)](https://search-proquest-com.ezproxy.lib.umb.edu/docview/1113980959?accountid=28932&clientId=46815&pq-origsite=summon) from nutrient enrichment affects a foundational species of nekton biodiversity that is interconnected to all of the creeks regardless of direct proximity to enrichment.

```{r interest_model, echo=FALSE, warning=FALSE}
nektonxmacro %>% filter(species %in% c('palaemonetes_pugio', 'fundulus_heteroclitus', 'streblospio_benedicti', 'manayunkia_aestuarina', 'menidia_menidia')) %>% ggplot(aes(x = date, y = count_percent, color = species)) +
  geom_point() +
  theme_minimal() +
  stat_smooth(se = FALSE, formula = 'y ~ x', method = 'loess') +
  facet_wrap(creek~.)
```
* There appears to be an unusual loss of equilibrium with observed creek sediment edge habitating worms in Nelson, the cause unclear in literature.
```{r contrast_species, echo=FALSE, warning=FALSE}
nmd_interest <- nektonxmacro %>% filter(species %in% c('palaemonetes_pugio', 'fundulus_heteroclitus', 'menidia_menidia'))
nek_spec_mod <- glm(count_percent ~ date + creek * species, data = nmd_interest, family=binomial(logit))
contrast(emmeans(nek_spec_mod, ~ creek * species))
```

|   A likelihood binomial logit regression model reveals some differences between species across the creeks, but with poor significance and standard error values it is difficult to confirm the potency of the effect. The regression model in itself is questionable with a tight resids vs fitted plot, a qq plot that has a fit center but tapering ends, yet small cook's distance values generally under 0.005. Poor significance of contrast related to the shrimp and silverback species is likely due to their low abundance in observation, a sample size error combined with the not quite fit model. However when looking at count percent data for the three nekton species specifically, the data 'appears' uniform but not normal based on skew/kurtosis values.

```{r contrast_model_check, echo=FALSE, warning=FALSE}
par(mfrow=c(1,3))
plot(nek_spec_mod, which = c(1, 2, 4))
```
```{r descdist, echo=FALSE, warning=FALSE}
descdist(nmd_interest$count_percent, graph = FALSE)
```
|    Returning to the literature there is an obvious effect on mummichog size because of geomorpological and bottom-up community change ([James A. Nelson et al.](https://link-springer-com.ezproxy.lib.umb.edu/article/10.1007/s10021-018-0265-x)) and change in its population abundance is clear in the model. Maybe then the lack of clarity or signifcance detected should be understood by exploring bottom-up effects specifically on the other species of interest in the model, silverbacks and the pugio shrimp rather then just trying to understand ecosystem diversity or population counts. Growing population size as shown in my model doesn't reflect the reality of poorer diets leading to smaller sized individuals. Looking at the literature and understanding the change in mummichog diet from not accessing the grasses I would deduce that silverbacks are more prevalent in their food search which would be a variable in the decreased silverback population; however, I wouldn't attribute silverback decline to that one relationship entirely. Without a firm understanding of new ecosystem relationships, risk effects, and predation due to a loss of a food source understanding treated areas is a game of analytics and not ecological fundamentals with my dataset. Abstracting it is clear that nekton and macroinfaunal species are declining more in treatment zones than control zones, but significance when parsing to that level is generally lost.

 \

```{r spec_model, echo=FALSE, warning=FALSE}
spec_model <- glm(total_count ~ date + branch * creek * type, data=nektonxmacro)
contrast(emmeans(spec_model, ~ branch * creek | type))
```
```{r general_model, echo=FALSE, warning=FALSE}
general_model <- glm(total_count ~ date + creek * type, data=nektonxmacro)
contrast(emmeans(general_model, ~ creek | type))
```
|    I think if I wanted to have a better understanding of population counts and the effects of treatment on them, I would need a more in depth framework with more specific covariates to interpret, a method to account for general anthropogenic noise or effects on all creeks, and a better understanding of modified individual behaviour in response to treatment. Optimally my guess is to understand nekton population representation I will need bayesian beta regression with a larger dataset. Without even having more behavioural data, I think understanding at what probability under what conditions I may pull a mummichog out of a hat will help with trying to create auxillary hypotheses about new top-down and bottom-up effects. 
